---
title: "Longitudinal data analysis: Hands-on in R"
author: "Tengku Muhammad Hanis Bin Tengku Mokhtar, PhD"
format: 
  html:
    toc: true
    toc-location: left
    theme: united
editor: visual
---

## Data

This practice will utilise a logitudinal clinical study data. The data consist of 60 subjects, 30 subjects were given drug A and another 30 subjects were given drug B. Each subjects were followed up for 6 weeks. The score measurements were taken each week.

The code below is used to create the data.

```{r}
# For reproducibility
set.seed(123)  

# Number of subjects and time intervals
n_subjects <- 60
n_drug_A <- n_subjects / 2
n_drug_B <- n_subjects / 2
n_time_intervals <- 6

# Generate the data
id <- rep(1:n_subjects, each = n_time_intervals)
drug <- rep(c(rep("A", n_time_intervals * n_drug_A), rep("B", n_time_intervals * n_drug_B)))
time <- rep(1:n_time_intervals, times = n_subjects)

# Define mean pulse scores for each group and time to create significant differences
# Time effect: increasing pulse over time
base_pulse_A <- seq(60, 70, length.out = n_time_intervals)
base_pulse_B <- seq(70, 80, length.out = n_time_intervals)

# Generate pulse scores with some random noise
score <- c(
  rep(base_pulse_A, times = n_drug_A) + rnorm(n_drug_A * n_time_intervals, mean = 0, sd = 2),
  rep(base_pulse_B, times = n_drug_B) + rnorm(n_drug_B * n_time_intervals, mean = 0, sd = 2)
)

# Create the data frame
df <- data.frame(id = factor(id), drug = factor(drug), week = factor(time), score = score)
```

## Packages

The packages below should be loaded.

```{r warning=FALSE, message=FALSE}
library(afex)
library(nlme)
library(emmeans)
library(tidyverse)
```

## Explore the data

First, we plot the boxplot.

```{r}
df %>% 
  ggplot(aes(x = week, y = score, fill = drug)) +
  geom_boxplot() +
  theme_bw()
```

## Multilevel model

### Develop the models

To run multilevel model, we need to create:

1.  Base or null model.
2.  Model with time variable (week).
3.  Model with time and group variable (week and drug).
4.  Model with time, group, and interaction between both both variables (week, group and week x group).

```{r}
# Base/null model
base <- lme(score ~ 1, random = ~1 | id/week, data = df, method = "ML")

# Add time variable - week
mod_week <- update(base, .~. + week) 

# Add group variable - drug
mod_week_drug <- update(mod_week, .~. + drug)

# Add interaction - group effect regardless of time
mod_int <- update(mod_week_drug, .~. + drug:week)
```

### Model comparison

Once we have all four models, we can do model comparison using AIC, BIC and likelihood ratio (LR) test.

```{r}
anova(base, mod_week, mod_week_drug, mod_int, type = "LR")
```

`mod_week_drug` has the lowest AIC and BIC. Also, LR test is not significant for the interaction. Thus, the best model is `mod_week_drug`.

### Assumption

For multilevel model, the only assumption is the normality of the residual.

```{r}
# Shapiro-wilk
shapiro.test(mod_week_drug$residuals)

# Kolmogorov-smirnov
DescTools::LillieTest(mod_week_drug$residuals) 
```

Both normality tests are significant (p \< 0.05).

```{r}
niceFunction::histWithCurve(mod_week_drug$residuals)  
```

However, using histogram we can see that the data seems to be quite normal-ish.

### Interpret

Lastly, we can interpret the final model.

```{r}
parameters::model_parameters(mod_week_drug)
```

We can see that drug B has a significant higher score than drug A (p \< 0.001). Also, each week comparison to week 1, all are significant (p \< 0.001).

Additionally, we can plot the marginal means.

```{r}
modelbased::estimate_means(mod_week_drug) %>%
  plot(show_data = "violin") +
  theme_bw()
```

## Repeated measure ANOVA

### Run repeated measure ANOVA

We can run repeated measure (RM) ANOVA using the following codes.

```{r message=FALSE, warning=FALSE}
mod <- aov_ez(id = "id", 
              dv = "score",
              between = "drug",
              within = "week",
              data = df)
```

Before we check the result, we need to check the assumptions.

### Assumptions

RM ANOVA has 3 assumptions:

1.  Normality of residuals.
2.  Equal variance.
3.  Sphericity.

Before we test the assumptions, we need to get the predicted and residual values.

```{r}
dat <- 
  data.frame(
    res = resid(mod),
    pred = predict(mod)
  )
```

1.  Normality test

```{r}
# Shapiro-wilk
shapiro.test(dat$res)

# Kolmogorov-smirnov
DescTools::LillieTest(dat$res) 

# Histogram
niceFunction::histWithCurve(dat$res)
```

The residuals are normally distributed.

2.  Equal variance.

```{r}
# Levene test
car::leveneTest(score ~ drug, data = df)

# Residual vs predicted values
dat %>% 
  ggplot(aes(x = pred, y = res)) +
  geom_point()

```

Equal variance assumption is met.

3.  Sphericity.

Sphericity assumption reflects equality of variances of the differences between all possible pairs of within-subject conditions. If this assumption is not met:

-   Mauchly's W \>0.75, take Huynh-Feldt correction.
-   Mauchly's W ≤0.75, take Greenhouse-Geisser correction.

```{r}

# 3) Mauchly's test (sphericity assumption)
summary(mod)$sphericity.tests
```

In our case, the sphericity assumption is met.

### Interpret

Next, we can interpret the result.

```{r}
summary(mod)
```

We can see that there is a significant difference of the score between the drug and week. However, the interaction between both variables is not significant.

### Post hoc tests

Since both variable drug and week is significant, we need to do a post hoc tests.

```{r}
# For drug
emmeans(mod, pairwise ~ drug)

# For week
emmeans(mod, pairwise ~ week, adjust = "bonferroni")
```
